Semantic Linearity in a Lazy Optimising Compiler

### 2025-10-07 popl26-accepted-before-rebuttal.tex

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
           \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
    \quad
            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \and \D'' = \D' \setminus (\Theta \setminus \Theta')}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma,
    \Sigma }
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Instrumented Operational Semantics~\label{fig:linopsem}}

\end{figure}

\subsection{Type Safety}\label{sec:typesafe}

Our results rely on a series of lemmas relating linear, unrestricted and
$\Delta$-bound variables, as well as a lemma characterizing irrelevant
resources.
%
% Additionally, we state our assumptions that outline an isomorphism between
% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
% resources $\D$, for any $\D$.

\renewcommand{\DeltaLinearRelationLemma}{
\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{restatable}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
\end{restatable}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{restatable}
}

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).

\LinearDeltaRelationLemma

\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
environment can be made into a linear variable as long as its
(irrelevant) environment is dropped from the linear context.

% Dually, a well-typed program with resources $\D$ and $\D$-bound
% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
% are removed from the linear context, and occurrences of $\D$ whole in usage
% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
% environments are unimportant since $\D$ was consumed whole by $x$, not by any
% of the fragment-using $\D$-vars).

\DeltaLinearRelationLemma

% Intuitively, a linear variable

\noindent Finally, unrestricted and $\D$-bound
variables with an empty usage environment are equivalent.

\DeltaUnrestrictedRelationLemma

\paragraph{Irrelevance}
%
Proof irrelevant resources can only be consumed
indirectly, essentially encoding that the (non-WHNF) case scrutinee
resources must be consumed through the case binder or the linear
pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually
reduce to WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, the two case rules must be in harmony in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.

% The type preservation theorem states that a well-typed expression
% remains well-typed in the presence of evaluation. Specifically, when the case
% expression whose scrutinee is evaluated to WHNF is handled in the
% preservation proof.

% The \emph{Irrelevance} lemma is required to prove preservation for that
% evaluation case. We need to prove that the alternatives of a case expression
% typed with proof irrelevant resources are still well-typed when the proof
% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
% case alternative with proof irrelevant resources in a certain context with respect to
% typing the same expression with arbitrary resources (we note, however, typing
% an alternative with proof irrelevant resources is not complete wrt using
% arbitrary resources -- a counter example needs only to use a resource
% directly).

\WHNFConvSoundness

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources with respect to
typing the same expression with arbitrary resources. However, typing
a case alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly.

% \noindent Intuitively, the lemma holds since proof irrelevant resources can
% only be used through the case binder or pattern-bound variables. If we
% consistently replace the proof irrelevant resources both in the typing
% environment and in the usage environments containing them, the expression
% remains well-typed. We note that the proof irrelevant resources are always
% unique when introduced in such a case alternative (we always take the scrutinee
% environment to make irrelevant, and allow nested ``irrelevantness''), so the
% case binder has the only occurrence of those resources in the $\G$ environment.
% % (being somewhat akin to congruence).
% %
% The proof is given in Section~\ref{sec:proof:irrelevance}.

%\subsection{Type safety\label{sec:type-safety-meta}}
% The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
% get \emph{stuck} and their evaluation preserves linearity.
We prove type safety of the Linear Core system via the standard type
preservation and progress results. As is customary, we make use of multiple
substitution lemmas, one for each kind of variable: unrestricted variables
$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
development is reported in Appendices~\ref{app:proofs}
and~\ref{app:opsem}, where type safety is established directly for the
instrumented semantics and ported to the natural semantics via a
bisimulation argument.

Note that since the operational semantics gets stuck if a linear
variable is used more than once, progress ensures that linear
functions consume their arguments at most once if their result is
consumed exactly once. Type preservation ensures that evaluation of a
closed term returns an evaluation environment with no linear variables.

  \begin{theorem}[Type Preservation]
    For any well-typed evaluation state $a$, if $a \Downarrow
    b$ or $a \Downarrow^* b$ then $b$ is well-typed.
  \end{theorem}

\begin{theorem}[Progress]
Let $a$ be a well-typed evaluation state. For any partial evaluation $a \Downarrow^* b$ 
the evaluation can be extended.
\end{theorem}

% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.

% \TypePreservationTheorem
% %
% \ProgressTheorem
% \noindent Type preservation states that a well-typed expression $e$ that
% evaluates to $e'$ remains well-typed under the same context:
%
% Type preservation
% % ($\S$~\ref{sec:proof:type-preservation})
% follows by structural induction on the reductions $e \longrightarrow
% e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
% substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
% interesting case is that of case expressions whose scrutinee can be further
% evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
%
% This case guarantees that the separation of rules for treating scrutinees is
% consistent, in the sense that a well-typed case expression with a scrutinee not
% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
%
%
%
%\noindent
% Progress states that the evaluation of a well-typed term does not block:
% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).

% \subsubsection{Substitution Lemmas}

% The preservation and progress theorems depend on multiple substitution lemmas,
% one for each kind of variable, as is standard.
% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
% % The ... themselves diverge from their common formulation, because
% % $\D$-variables refer to linearly bound variables, so substitution must take into account

% The linear substitution lemma states that a well-typed expression $e$ with a
% linear variable $x$ of type $\s$ remains well-typed if
% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
% type $\s$, and occurrences of $x$ in the linear context and in usage
% environments of $\D$-bound variables are replaced by the linear context $\D'$
% used to type $e'$:

% \LinearSubstitutionLemma

% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).

% The substitution of the resource in the usage environments is illustrated
% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
% the usage environment of $y$ accordingly, we'll ultimately be typing
% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.

% The linear substitution lemma extends to case alternatives as well.
% The lemma for substitution of linear variables in case alternatives is similar
% to the linear substitution lemma, applied to the case alternative judgement.
% % Is slightly more involved, in the sense that there are more environments in which the
% % substitution $[\D/x]$ must be applied (for the same reason):
% %
% \LinearSubstitutionAltsLemma
% %
% \noindent We further require that the environment annotated in the case
% alternative judgement, $\D_s$, is a subset of the environment used to type the
% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
% environment annotating the alternative judgement is \emph{always} a subset of
% the alternative environment.

% The substitution lemma for unrestricted variables follows the usual
% formulation, with the added restriction (common to linear type systems) that
% the expression $e'$ that is going to substitute the unrestricted variable $x$
% is typed on an empty linear environment:
% %
% \UnrestrictedSubstitutionLemma
% %
% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
% %
% \UnrestrictedSubstitutionAltsLemma

% Finally, we introduce the lemma stating that substitution of $\D$-bound
% variables by expressions of the same type preserves the type of the original
% expression.
% %
% What distinguishes this lemma from traditional substitution lemmas is that the
% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
% must match exactly the typing environment $\D$ of $e'$ and the
% environment of the original expression doesn't change with the substitution:
% %
% \DeltaSubstitutionLemma
% %
% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
% by an expression $e'$ which is typed in the same environment $\D$ allows the
% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
% substitution lemmas), as we only ever substitute $\D$-variables by expressions
% whose typing environment matches the variables usage environment. However, it
% is not obvious whether such a lemma is possible to prove for $\D$-variables
% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).

% The $\D$-substitution lemma on case alternatives reflects again that the typing
% environment of the expression substitution the variable must match its usage
% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
% environment is always contained in the typing environment, which is true of all
% occurrences of this judgement. An alternative formulation of this lemma could
% instead explicitly list $\D_s$ as part of the typing environment for the same
% effect:

% \DeltaSubstitutionAltsLemma
